Synthesis of Linear Functional Programs

### 2021-07-23 article/article.tex
Table and almost done

\mypara{Core Rules} The system comprises of proof search in
(intuitionistic) linear logic sequent calculus, based on a system of
resource management~\cite{resman} and focusing.

@@

The semantics of ADTs relate to those of the plus ($\oplus$) type -- both are
additive disjunctions.  To construct a value of an ADT we must use one of its
constructors, similar to the way $\oplus$ requires only proof of either the left
or right type it consists of. Analogously, we can deconstruct a value of an ADT
via pattern matching on its constructors, where all branches of the pattern
match must have the same type -- akin to the left rule for the $\oplus$
connective. In effect, the inference rules for a simple ADT are a generalized
form of the $\oplus$ rules.  Therefore, there's one left rule for ADTs, and an
arbitrary number of right rules, one for each constructor, where ADT $T$ and
its constructors stand for any ADT defined as \mintinline{haskell}{data T = C1
X1 | C2 X2 | ... | Cn Xn}
%
\begin{mathpar}
    \infer*[right=(adtR)]
    {\Gamma; \Delta/\Delta' \vdash M : X_n \Downarrow}
    {\Gamma; \Delta/\Delta' \vdash\ C_n \ M : T \Downarrow}
\and
    \mprset{flushleft}
    \infer*[right=(adtL)]
    {
        \Gamma ; \Delta/ \Delta'_1 ; \Omega, y_1{:}X_1 \Uparrow\ \vdash M_1 : C
        \and
        y_1 \notin \Delta'_1
        \\
        \Gamma ; \Delta/ \Delta'_2 ; \Omega, y_2{:}X_2 \Uparrow\ \vdash M_2 : C
        \\
        y_2 \notin \Delta'_2
        \\\\
        \\ \dots
        \\
        \Gamma ; \Delta/ \Delta'_n ; \Omega, y_n{:}X_n \Uparrow\ \vdash M_n : C
        \\
        y_n \notin \Delta'_n
        \\
        \Delta'_1 = \Delta'_2 = \dots = \Delta'_n
    }
    {\Gamma ; \Delta/\Delta'_1 ; \Omega, x{:}T \Uparrow\
    \vdash\ \textrm{case}\ x\ \textrm{of}\ \dots\ |\ C_n\ y_n
    \rightarrow M_n : C}
\end{mathpar}

@@

% TODO TODO: Big rewrite
The other situation occurs when using an unrestricted polymorphic function that
requires synthesis of a term with an existential type when the goal is an
existential type. In contrast to the previous problem, the type of the goal and
of the argument that will cause the loop won't match exactly, since instantiated
bound variables are always fresh. For example, for $\forall \alpha,\beta .\
\alpha\lolli\beta\lolli\beta;?\alpha\lolli?\beta\lolli?\beta \Downarrow\ \vdash\
?\sigma$, we'll unify $?\beta$ with $?\sigma$, and then require a term of type
$?\beta$ (not $?\sigma$). We want to forbid the usage of the \emph{same}
function to attain \emph{any} existential goal, provided that function might
create existential sub-goals (i.e. it's polymorphic). However, we noticed that,
even though for most tried problems this ``same function'' approach worked,
context-heavy problems such as \emph{array} (seen in Section~\ref{sec:overview})
wouldn't terminate in a reasonable amount of time.
%
% In fact, with $n$ polymorphic functions in the unrestricted context, the
% complexity of searching for a program of any existential type $?\alpha$, while
% restricting solely the used function, is $O(n!)$~\ref{exemplo_apendice?}.
%
As such, we'll instead forbid the usage of \emph{all} unrestricted polymorphic
functions to synthesize any existential type if, altogether, we've already used
$d_e$ times a polymorphic function for \emph{any} existential sub-goal, where
$d_e$ is a \emph{constant existential depth} that controls the maximum depth of
``existential synthesis''. \todo{alinhar mais com a premissa
  respectiva da regra}
%
% This approach reduces the complexity of our proof-search algorithm for
% existential types to $O(\tfrac{n!}{d_e!}) = O(n^{d_e})$.

Extending the restrictions context ($\Rho$) with restrictions on using the
unrestricted context ($\Rho_{L!}$), we modify \textsc{decideLeft!} to formalize
the previous paragraph:
\begin{mathpar}
    \mprset{flushleft}
    \infer*[right=(decideL!)]
    {
    (A, C) \notin \Rho_{L!}
    \\
    \textsc{isExist}(C) \Rightarrow |\{u\ |\
    (f,u)\in\Rho_{L!},\textsc{isExist}(u),\textsc{isPoly}(f)\}| < d_e
    \\
    \Theta/\Theta';(\Rho_C,\Rho_D,\Rho_{L!}');\Gamma, A; \Delta/\Delta' ; A
    \Downarrow\ \vdash C
    }
    {\Theta/\Theta';(\Rho_C,\Rho_D,\Rho_{L!});\Gamma, A; \Delta/\Delta';\cdot \Uparrow\ \vdash C}
    \\
    \Rho_{L!}' =\ \textbf{if}\ \textsc{isFunction}(A)\ \textbf{then}\ \Rho_{L!},(A, C)\ \textbf{else}\ \Rho_{L!}
\end{mathpar}

@@

\mypara{Refinement Types} Refinement types are defined as a types
with a name and with a predicate (a non-existing predicate is the same as it
being always \emph{true}); dependent types are functions with
refinement types in which the argument type name can be used in the predicates
of the return type (e.g. $x\{\mathsf{Int}\} \lolli y\{\mathsf{Int}\ \vert\ y = x\}$ specifies a function that
takes an Int named $x$ and produces an Int whose value is equal to the input $x$). We extend
the types syntax with our refinement types syntax:
\[
\begin{tabular}{lclc}
    $\tau$ & $\ ::=\ $ & $ \dots \vert\  (x:\tau) \lolli \sigma\ \vert\  \{x:\tau\ \vert\ P\}$ \\
    $P$ & $\ ::=\ $ & $P = P\ \vert\ P \neq P\ \vert\ P \vee P\ \vert\ P \wedge
    P\ \vert\ P \Rightarrow P\ \vert\ n = n\ \vert\ n \neq n$ \\
    & & $\vert\ n \leq n\ \vert\ n \geq n\ \vert\ n < n\ \vert\ n > n\ \vert\
    true\ \vert\ false\ \vert\ x$ \\
    $n$ & $\ ::=\ $ & $n * n\ \vert\ n + n\ \vert\ n - n\ \vert\
    \langle\emph{natural}\rangle\ \vert\ x$
    \\
\end{tabular}
\]
The addition of refinement types to the synthesizer doesn't
interfere with the rest of the process. We define the following right and
left rule, to synthesize or consume in synthesis a refinement type, where
$\textsc{getModel}(p)$ is a call to an SMT solver that returns a model of an
uninterpreted function that satisfies
$\forall_{\tilde{1},\tilde{2},\dots,\tilde{n}}\ h_{\tilde{1}} \Rightarrow h_{\tilde{2}} \Rightarrow \dots
\Rightarrow h_{\tilde{n}} \Rightarrow p$, where $\tilde{i}$ is the refinement type
name and $h_{\tilde{i}}$ its predicate for every refinement type in the
propositional contexts; and $\textsc{sat}(p_{\tilde{y}} \Rightarrow p_{\tilde{x}})$ is a call to an SMT solver that
determines universal satisfiability of the implication between predicates (the
left focused proposition subtypes the goal).
\begin{mathpar}
    \infer*[right=(refR)]
    { \textsc{getModel}(p) = M }
    {\Theta/\Theta';\Rho;\Gamma;\Delta \vdash M : \tilde{x}\{A\ \vert\
    p\}\Uparrow }
    \and
    \infer*[right=(refL)]
    { \textsc{sat}(p_{\tilde{y}} \Rightarrow p_{\tilde{x}}) }
    {\Theta/\Theta';\Rho;\Gamma;\Delta; x{:}\tilde{y}\{A\ \vert\ p_{\tilde{y}}\}\Downarrow\ \vdash x : \tilde{x}\{A\ \vert\ p_{\tilde{x}}\} }
\end{mathpar}

@@

The desugaring module converts the frontend AST to a core AST without
syntatic sugar and using the locally-nameless
representation~\cite{locally nameless}, in which all bound variables
are represented with Debruijn indices rather than names (i.e.~using
the distance to the binding site in terms of the number of traversed
binders),
% TODO: não ficou muito mais claro :\todo{Explica o que e isto}
This way inference can be done without worrying about name conflicts.

The inference module will traverse the list of ASTs in order (top-down), and
infer the type of all functions defined by the user, checking the inferred
type against a possible user given type annotation. Whenever a synth mark is
found in the AST, all the inferrence context up to that point is added to the
mark, and if not explicit, its type is inferred, essentially defining the synth
context and synth goal.

The synthesis module iterates over the synthesis marks and runs the synthesizer
with the respective context and goal for each mark. The resulting expression is checked
against the synthesis constraints defined through some keywords, continuing
synthesis if they are violated: ``using'' guarantees that certain functions are
used in the program body, and ``assert'' evaluates the given assertion within the
synthesized program to a validating boolean.  Additionally, the keyword
``depth'' controls the \emph{existential depth} (seen
in~\ref{sec:formal_system}), and ``choose'' selects a different result as long
as one is available. Finally, the expression is minimized (i.e.~$\eta$-reduced),
and the mark in the frontend AST is replaced with the synthesized program.

The evaluation module evaluates the function named ``main'' within the
complete synthesized program, returning a final value.

@@

@@

\mypara{Interface} A simple, syntax-highlighted, web interface was developed
alongside a server to make possible experimenting with the \synname\ 
language without having to download and compile the complete toolchain. The web
interface is available from the repository~\cite{github repo}.

@@

Type-based program synthesis is a vast field of study and so it follows that a
lot of literature is available to inspire and complement our work. Most
works~\cite{DBLP:conf/lopstr/HughesO20,DBLP:conf/pldi/PolikarpovaKS16,DBLP:conf/pldi/OseraZ15,DBLP:conf/popl/FrankleOWZ16}
follow some variation of the synthesis-as-proof-search approach. However, the
process is novel for each due to a variety of different rich types explored and
their corresponding logics and languages; or nuances of the synthesis process
itself, such as complementing types with program examples; or even the
programming paradigm of the output produced (e.g. generating heap manipulating
programs~\cite{DBLP:journals/pacmpl/PolikarpovaS19}).  Some other common
patterns we follow are the use of type
refinements~\cite{DBLP:conf/pldi/PolikarpovaKS16}, polymorphic types, and the
support for synthesis of recursive
functions~\cite{DBLP:conf/pldi/PolikarpovaKS16,DBLP:conf/pldi/OseraZ15}. These
works share some common ground with each other and with us -- most
famously the Curry-Howard correspondence, but also, for example, 
focusing-based proof-search~\cite{10.1093/logcom/2.3.297}.
% TODO: sinto que precisava de um toque...

@@

\mypara{Program Synthesis from Polymorphic Refinement Types} The
work~\cite{DBLP:conf/pldi/PolikarpovaKS16} also studies synthesis of recursive
functional programs in an ``advanced'' context. Their specifications combine two
rich forms of types: polymorphic and refinement types (which correspond to a
first-order logic through the Curry-Howard isomorphism). Their approach to
refinement types consists of a new algorithm that supports decomposition of the
refinement specification, allowing for separation between the language of
specification and programs and making the approach amenable to compositional
synthesis techniques. We also support refinements (and polymorphism),
but they are not as integrated in the synthesis process as in \cite{DBLP:conf/pldi/PolikarpovaKS16}.
Instead, our synthesizer leverages the expressiveness of linear
types and techniques for proof-search in linear logic to guide its process.

\mypara{Resourceful Program Synthesis from Graded Linear Types} The
work~\cite{DBLP:conf/lopstr/HughesO20} synthesizes programs using an
approach very similar to our own.  The work employs so-called graded
modal types, which is a refinement of pure linear types -- a more
\emph{fine-grained} version, since it allows for quantitative
specification of resource usage, in contrast to ours either
\emph{linear} or \emph{unrestricted} (via the linear logic
exponential) use of assumptions.  Their resource management is more
complex, and so they provide solutions which adapt Hodas and Miller's
resource management
approach~\cite{DBLP:journals/tcs/CervesatoHP00,DBLP:journals/tcs/LiangM09}
-- which, in contrast, is the model use in our work.

They also use focusing as a solution to trim down search space and to
ensure that synthesis only produces well-typed programs. However, since their
underlying logic is \emph{modal} rather than purely \emph{linear}, it
lacks a clear correspondence with concurrent session-typed
programs~\cite{DBLP:journals/mscs/CairesPT16,DBLP:conf/concur/CairesP10},
which is a crucial avenue of future work. Moreover, their use grading
effectively requires constraint solving to be integrated with the
synthesis procedure, which can limit the effectiveness of the overall
approach as one scales to more sophisticated settings (e.g.~refinement
types). \todo{but we have polymorphism and recursion}

@@

The benchmarks are separated into groups: theorems of linear logic, recursive
ADTs, refinements. For each goal we analyze if the synthesis process terminated
according to what was expected, the mean time and standard deviation, if the
resulting program fullfills its intended purpose

\begin{figure}[t]
{\scriptsize
\begin{center}
    \begin{tabular}{ |c|c|c|c|c| }
        \hline
        Group & Goal & Avg. time + $\sigma$ & Found & Is expected \\
        \hline
        \multirow{4}{4em}{Linear Logic Theor.} & uncurry & 1s & yes & yes \\ 
        & curry & 2s & yes & yes \\ 
        & call by name?? & 2s & yes & yes \\ 
        & 0/1?? & 2s & yes & yes \\ 
        \hline
        \multirow{8}{4em}{List} & singleton & 3s & yes & yes \\ 
        & insert & 2s & yes & unordered \\ 
        & map & 1s & yes & yes \\ 
        & append & 2s & yes & yes \\ 
        & foldl & 2s & yes & got (foldl . reverse) \\ 
        & foldr & 2s & yes & yes \\ 
        & concat & $\infty$ & no & no \\ 
        & uncons & 1s & yes & yes \\
        \hline
        \multirow{3}{4em}{Maybe} & return & 1s & yes & yes \\ 
        & $>>=$ & 1s & yes & yes \\
        & maybe & 1s & yes & yes \\
        & list to maybe & 12s & no & yes \\
        \hline
        \multirow{5}{4em}{Binary Tree} & singleton & 1s & yes & yes \\ 
        & insert & 1s & yes & unordered \\
        & map & 1s & yes & yes \\
        & append & 1s & yes & appends to rightmost leaf \\
        & join node & 1s & yes & joins at rightmost leaf \\
        \hline
        \multirow{7}{4em}{Map} & singleton & 1s & yes & yes \\ 
        & insert & 1s & yes & unordered \\
        & insertWithKey & 1s & yes & ignores function \\
        & union & 1s & yes & ignores keys \\
        & mapAccum & 1s & no & ?? \\
        & map & 1s & yes & yes \\
        & mapKeys & 1s & yes & yes \\
        \hline
        \multirow{1}{4em}{Misc} & either & 1s & yes & yes \\ 
        \hline
    \end{tabular}
  \end{center}
}
\caption{Table with \dots}
\end{figure}

Synthesis of

>>>>

\mypara{Core Rules} The system comprises of proof search in
(intuitionistic) linear logic sequent calculus, based on a system of
resource management~\cite{DBLP:journals/tcs/CervesatoHP00,DBLP:journals/tcs/LiangM09}
and focusing.

@@

The semantics of ADTs relate to those of the plus ($\oplus$) type -- both are
additive disjunctions.  To construct a value of an ADT we must use one of its
constructors, similar to the way $\oplus$ requires only proof of either the left
or right type it consists of. Analogously, we can deconstruct a value of an ADT
via pattern matching on its constructors, where all branches of the pattern
match must have the same type -- akin to the left rule for the $\oplus$
connective. In effect, the inference rules for a simple ADT are a generalized
form of the $\oplus$ rules.  Therefore, there's one left rule for ADTs, and an
arbitrary number of right rules, one for each constructor, where ADT $T$ and
its constructors stand for any ADT defined as \mintinline{haskell}{data T = C1
X1 | C2 X2 | ... | Cn Xn}:
%
\begin{mathpar}
    \infer*[right=(adtR)]
    {\Gamma; \Delta/\Delta' \vdash M : X_n \Downarrow}
    {\Gamma; \Delta/\Delta' \vdash\ C_n \ M : T \Downarrow}
\and
    \mprset{flushleft}
    \infer*[right=(adtL)]
    {
        \Gamma ; \Delta/ \Delta'_1 ; \Omega, y_1{:}X_1 \Uparrow\ \vdash M_1 : C
        \and
        y_1 \notin \Delta'_1
        \\
        \Gamma ; \Delta/ \Delta'_2 ; \Omega, y_2{:}X_2 \Uparrow\ \vdash M_2 : C
        \\
        y_2 \notin \Delta'_2
        \\\\
        \\ \dots
        \\
        \Gamma ; \Delta/ \Delta'_n ; \Omega, y_n{:}X_n \Uparrow\ \vdash M_n : C
        \\
        y_n \notin \Delta'_n
        \\
        \Delta'_1 = \Delta'_2 = \dots = \Delta'_n
    }
    {\Gamma ; \Delta/\Delta'_1 ; \Omega, x{:}T \Uparrow\
    \vdash\ \textrm{case}\ x\ \textrm{of}\ \dots\ |\ C_n\ y_n
    \rightarrow M_n : C}
\end{mathpar}

@@

The other situation occurs when using an unrestricted polymorphic function that
requires synthesis of a term with an existential type when the goal is an
existential type. In contrast to the previous problem, the type of the goal and
of the argument that will cause the loop won't match exactly, since instantiated
bound variables are always fresh. For example, for $\forall \alpha,\beta .\
\alpha\lolli\beta\lolli\beta;?\alpha\lolli?\beta\lolli?\beta \Downarrow\ \vdash\
?\sigma$, we'll unify $?\beta$ with $?\sigma$, and then require a term of type
$?\beta$ (not $?\sigma$). We want to forbid the usage of the \emph{same}
function to attain \emph{any} existential goal, provided that function might
create existential sub-goals (i.e. it's polymorphic). However, we noticed that,
even though for most tried problems this ``same function'' approach worked,
context-heavy problems such as \emph{array} (seen in \S~\ref{sec:overview})
wouldn't terminate in a reasonable amount of time.
%
% In fact, with $n$ polymorphic functions in the unrestricted context, the
% complexity of searching for a program of any existential type $?\alpha$, while
% restricting solely the used function, is $O(n!)$~\ref{exemplo_apendice?}.
%
As such, we'll instead define that, given an existential\footnote{a type 
is existential when any of its components is an existential type variable} goal $C$,
we can only ``decide left!'' on a proposition $A$ if, altogether, 
the amount of times we've ``decided left!'' on an polymorphic function to produce
an existential goal is less than a \emph{constant ``existential depth''} $d_e$
(which controls a \emph{depth} aspect of the synthesis process).
%
% This approach reduces the complexity of our proof-search algorithm for
% existential types to $O(\tfrac{n!}{d_e!}) = O(n^{d_e})$.

Extending the restrictions context ($\Rho$) with restrictions on using the
unrestricted context ($\Rho_{L!}$), we modify \textsc{decideLeft!} to formalize
the two previous paragraphs, where $\textsc{isExist}(C)$ is true if $C$ is an
existential type, $\textsc{isPoly}(f)$ is true if $f$ is universally
quantified (i.e. $f$ has form $\forall\overline{\alpha} f'$), and
$\Rho_{L!}' = \Rho_{L!},(A, C)$ if $A$ is a function and $\Rho_{L!}' =
\Rho_{L!}$ otherwise:
%
\begin{mathpar}
    \mprset{flushleft}
    \infer*[right=(decideL!)]
    {
    (A, C) \notin \Rho_{L!}
    \\
    \textsc{isExist}(C) \Rightarrow |\{u\ |\
    (f,u)\in\Rho_{L!},\textsc{isPoly}(f),\textsc{isExist}(u)\}| < d_e
    \\
    \Theta/\Theta';(\Rho_C,\Rho_D,\Rho_{L!}');\Gamma, A; \Delta/\Delta' ; A
    \Downarrow\ \vdash C
    }
    {\Theta/\Theta';(\Rho_C,\Rho_D,\Rho_{L!});\Gamma, A; \Delta/\Delta';\cdot \Uparrow\ \vdash C}
    \\
\end{mathpar}

@@

\mypara{Refinement Types} Refinement types are types with a predicate (a non-existing predicate is the same as it
being \emph{true}); dependent types are functions with
refinement types in which the argument type is labeled and said label can be used in the predicates
of the return type (e.g. $(x : \mathsf{Int}) \lolli \{y : \mathsf{Int}\ \vert\ y = x\}$ specifies a function that
takes an Int and returns an Int of equal value). We extend
the types syntax with our refinement types:
\[
\begin{tabular}{lclc}
    $\tau$ & $\ ::=\ $ & $ \dots \vert\  (x:\tau) \lolli \sigma\ \vert\  \{x:\tau\ \vert\ P\}$ \\
    $P$ & $\ ::=\ $ & $P = P\ \vert\ P \neq P\ \vert\ P \vee P\ \vert\ P \wedge
    P\ \vert\ P \Rightarrow P\ \vert\ n = n\ \vert\ n \neq n$ \\
    & & $\vert\ n \leq n\ \vert\ n \geq n\ \vert\ n < n\ \vert\ n > n\ \vert\
    true\ \vert\ false\ \vert\ x$ \\
    $n$ & $\ ::=\ $ & $n * n\ \vert\ n + n\ \vert\ n - n\ \vert\
    \langle\emph{natural}\rangle\ \vert\ x$
    \\
\end{tabular}
\]
The addition of refinement types to the synthesizer doesn't
interfere with the rest of the process. We define the following right and
left rule, to synthesize or consume in synthesis a refinement type, where
$\textsc{getModel}(p)$ is a call to an SMT solver that returns a model of an
uninterpreted function that satisfies
$\forall_{a,b,\dots,n}\ h_{a} \Rightarrow h_{b} \Rightarrow \dots
\Rightarrow h_{n} \Rightarrow p$, where $n$ is the refinement type label
and $h_{n}$ its predicate, with $a,\dots,n$ standing for the label of every refinement type
in the propositional contexts;
and $\textsc{sat}(p_{a} \Rightarrow p_{b})$ is a call to an SMT solver that
determines universal satisfiability of the implication between predicates (the
left focused proposition subtypes the goal).
\begin{mathpar}
    \infer*[right=(refR)]
    { \textsc{getModel}(p) = M }
    {\Theta/\Theta';\Rho;\Gamma;\Delta/\Delta' \vdash M : \{a : A\ \vert\
    p\}\Uparrow }
    \and
    \infer*[right=(refL)]
    { \textsc{sat}(p_{a} \Rightarrow p_{b}) }
    {\Theta/\Theta';\Rho;\Gamma;\Delta/\Delta'; x{:}(\{a : A\ \vert\
    p_{a}\})\Downarrow\ \vdash x : \{b : A\ \vert\ p_{b}\} }
\end{mathpar}

@@

The desugaring module converts the frontend AST to a core AST without
syntatic sugar and using the locally-nameless
representation~\cite{locally nameless}, in which all bound variables
are represented with Debruijn indices rather than names (i.e.~using
the distance to the binding site in terms of the number of traversed
binders),
This way inference can be done without worrying about name conflicts.

The inference module will traverse the list of ASTs in order (top-down), and
infer the type of all functions defined by the user, checking the inferred
type against a possible user given type annotation. Whenever a synthesis mark is
found in the AST, all the inferrence context up to that point is added to the
mark, and, if not explicit, its type is inferred, essentially defining the
synthesis context and synthesis goal.

The synthesis module iterates over the marks and runs the synthesizer
with the respective context and goal for each mark. The resulting expression is checked
against the synthesis constraints defined through some keywords, continuing
synthesis if they are violated: ``using'' guarantees that certain functions are
used in the program body (to invalidate valid program without them), and ``assert'' evaluates the given assertion within the
synthesized program to a validating boolean.  Additionally, the keyword
``depth'' controls the \emph{existential depth} (seen
in \S~\ref{sec:formal_system}), and ``choose'' selects a different result as long
as one is available. Finally, the expression is minimized (i.e.~$\eta$-reduced),
and the mark in the frontend AST is replaced with the synthesized program.

The evaluation module evaluates the function named ``main'' within the
complete synthesized program, returning a final value. When evaluating
synthesis assertions, a timeout is used to interrupt possibly non-terminating
programs.

@@

The implementation was carried out in Haskell. We highlight some key points. The
project and source code are available from the repository~\cite{sili_repo}.

@@

\mypara{Interface} A simple, syntax-highlighted, web interface was developed
alongside a server to make possible experimenting with the \synname\ 
language without having to download and compile the complete toolchain. The web
interface is available from the repository~\cite{sili_repo}.

@@

Type-based program synthesis is a vast field of study and so it follows that a
lot of literature is available to inspire and complement our work. Most
works~\cite{DBLP:conf/lopstr/HughesO20,DBLP:conf/pldi/PolikarpovaKS16,DBLP:conf/pldi/OseraZ15,DBLP:conf/popl/FrankleOWZ16}
follow some variation of the synthesis-as-proof-search approach. However, the
process is novel for each due to a variety of different rich types explored and
their corresponding logics and languages; or nuances of the synthesis process
itself, such as complementing types with program examples; or even the
programming paradigm of the output produced (e.g. generating heap manipulating
programs~\cite{DBLP:journals/pacmpl/PolikarpovaS19}).
% Some other common
% patterns we follow are the use of type
% refinements~\cite{DBLP:conf/pldi/PolikarpovaKS16}, polymorphic types, and the
% support for synthesis of recursive
% functions~\cite{DBLP:conf/pldi/PolikarpovaKS16,DBLP:conf/pldi/OseraZ15}. These
% works share some common ground with each other and with us -- most
% famously the Curry-Howard correspondence, but also, for example, 
% focusing-based proof-search~\cite{10.1093/logcom/2.3.297}.

@@

\mypara{Program Synthesis from Polymorphic Refinement Types} The
work~\cite{DBLP:conf/pldi/PolikarpovaKS16} also studies synthesis of recursive
functional programs in an ``advanced'' context. Their specifications combine two
rich forms of types: polymorphic and refinement types.
% (which correspond to a
% first-order logic through the Curry-Howard isomorphism).
Their approach to
refinement types consists of a new algorithm that supports decomposition of the
refinement specification.
%,allowing for separation between the language of
%specification and programs and making the approach amenable to compositional
%synthesis techniques.
We also support refinements (and polymorphism),
but they are not as integrated in the synthesis process as in \cite{DBLP:conf/pldi/PolikarpovaKS16}.
Instead, our synthesizer leverages the expressiveness of linear
types and techniques for proof-search in linear logic to guide its process.

\mypara{Resourceful Program Synthesis from Graded Linear Types} The
work~\cite{DBLP:conf/lopstr/HughesO20} synthesizes programs using an
approach similar ours. It employs so-called graded
modal types, which are a refinement of pure linear types
% -- a more
% \emph{fine-grained} version 
that allows for quantitative
specification of resource usage, in contrast to ours either
\emph{linear} or \emph{unrestricted} (via the linear logic
exponential) use of assumptions. Their resource management is more
complex, and so they provide solutions which adapt Hodas and Miller's
approach~\cite{DBLP:journals/tcs/CervesatoHP00,DBLP:journals/tcs/LiangM09}.
% -- which, in contrast, is the one used in our work.
%
They also use focusing as a solution to trim down search space and to
ensure that synthesis only produces well-typed programs. However, since their
underlying logic is \emph{modal} rather than purely \emph{linear}, it
lacks a clear correspondence with concurrent session-typed
programs~\cite{DBLP:journals/mscs/CairesPT16,DBLP:conf/concur/CairesP10},
which is a crucial avenue of future work. Moreover, their use of grading
effectively requires constraint solving to be integrated with the
synthesis procedure, which can limit the effectiveness of the overall
approach.
% as one scales to more sophisticated settings (e.g.~refinement
% types).
Additionally, Our system extends the focusing-based system with
recursion, ADTs, polymorphism and refinements to synthesize
more expressive programs.

@@

The benchmarks\footnote{Considering the full pipeline pass; 2017 MBP; 8GB RAM;
2.3Ghz Intel Core i5; with benchmarking library Criterion} are separated into groups: theorems of linear logic, recursive
ADTs, refinements. The first table displays successful synthesis -- for each
goal we analyze the mean time and standard deviation, the conditions used to
guide the output, and the components (other functions) used in the resulting
program.

% TODO Verificar nomes dos linear logic theoresm...
{\scriptsize
\begin{center}
    \begin{tabular}{ |c|c|c|c|c| }
        \hline
        Group & Goal & Avg. time $\pm\,\sigma$ & Keywords & Components \\
        \hline
        \multirow{4}{4em}{Linear Logic Theor.} & uncurry & $133\mu s\pm 4.9\mu s$ &&\\ 
        & distributivity & $179\mu s\pm 5.0\mu s$ && \\ 
        & call by name & $196\mu s\pm 4.6\mu s$ && \\ 
        & 0/1 & $294\mu s\pm 5.3\mu s$ && \\ 
        \hline
        \multirow{7}{4em}{List} & map & $288\mu s\pm 7.2\mu s$  && \\ 
        & append & $292\mu s\pm 7.0\mu s$ & & \\ 
        & foldl & $1.69ms\pm 5.3\mu s$ & \emph{choose 1} & \\ 
        & foldr & $704\mu s\pm 10\mu s$ && \\ 
        & concat & $505\mu s\pm 18\mu s$ && \\
        & uncons & $215\mu s\pm 15\mu s$ && \\
        & reverse & $17.4ms\pm 515\mu s$ & \emph{reverse [1,2] == [2,1]} & \\ 
        \hline
        \multirow{2}{4em}{Maybe} & $>>=$ & $194\mu s\pm 5.3\mu s$ && \\ 
        & maybe & $161\mu s\pm 4.8\mu s$&& \\
        \hline
        \multirow{7}{4em}{State} & runState & $190\mu s\pm 6.8\mu s$ && \\ 
        & $>>=$ & $979\mu s\pm 23\mu s$ && \\
        & $>>=$ & $\infty$ & \emph{using (runState)} & \\
        & get & $133\mu s\pm 3.8\mu s$ && \\
        & put & $146\mu s\pm 3.4\mu s$ && \\
        & modify & $219\mu s\pm 4.9\mu s$ && \\
        & evalState & $156\mu s\pm 4.0\mu s$ && \\
        % \multirow{5}{4em}{Binary Tree} & singleton & 1s & yes & yes \\ 
        % & insert & 1s & yes & unordered \\
        % & map & 1s & yes & yes \\
        % & append & 1s & yes & appends to rightmost leaf \\
        % & join node & 1s & yes & joins at rightmost leaf \\
        % \hline
        % \multirow{7}{4em}{Map} & singleton & 1s & yes & yes \\ 
        % & insert & 1s & yes & unordered \\
        % & insertWithKey & 1s & yes & ignores function \\
        % & union & 1s & yes & ignores keys \\
        % & mapAccum & 1s & no & ?? \\
        % & map & 1s & yes & yes \\
        % & mapKeys & 1s & yes & yes \\
        \hline
        \multirow{1}{4em}{Misc} & either & $197\mu s\pm 5.3\mu s$ && \\ 
        \hline
        \multirow{2}{4em}{Array} & & & \emph{depth 3} & \emph{freeze, foldl} \\
        & array & $80ms\pm 870\mu s$ & \emph{using (foldl),depth 3} & \emph{newMArray,write} \\
        \hline
    \end{tabular}
  \end{center}
}
